A Kochen-Specker system has at least 21 vectors\\ {\small (extended abstract)}

### 2014-09-10 paper/main.tex
paper: *not* biconnected
paper: edge-{tri,bi}connected

We continue with a strengthening of Proposition~\ref{prop:ks-conn}.
\begin{prop}\label{prop:ks-biconn}
A minimal Kochen-Specker graph is biconnected,
that is: removing any single edge leaves the graph connected.
\end{prop}
We need some preparation, before we can prove this Proposition.
\begin{dfn}
Given a graph~$G$ and a vertex~$v$ of~$G$.
We say, \keyword{$v$ has fixed color~$c$ (in~$G$)},
if~$G$ is~$010$-colorable
and for every~$010$-coloring of~$G$,
the vertex~$v$ is assigned color~$c$.
\end{dfn}
We are interested in these graphs because of the following observation.
\begin{lem}\label{lem:biconn}
If there is an embeddable graph~$G$ on~$n$ vertices with a vertex
with fixed color~$1$,
then there is a Kochen-Specker graph on~$2n$ vertices.
\end{lem}
\begin{proof}
Let~$G$ be a graph and~$v$ a vertex of~$G$ with fixed color~$1$.
Consider two copies of the graph~$G$.
Connect the two instances of~$v$ with an edge.
Call this graph~$G'$.
Clearly, $G'$ is not~$010$-colorable.

@@

Suppose there are points in~$S'$ and~$S$
that overlap.
Note that the north pole (and south pole) is not in~$S'$. 
Let~$S''$ be~$S'$ rotated along the north pole at some angle~$\alpha$.
There are finitely many angles such that there are overlapping points.
Thus there is an angle such that~$S \cup S''$ is an embedding of~$G'$.
\end{proof}
Unfortunately, these graphs are not small.
\begin{comp}\label{comp:bic1}
Let~$F^{3,1}_n$ denote the number of connected graphs
with minimal vertex-order~$3$
on~$n$ vertices
with a vertex with fixed color~$1$.
Then:
\begin{center}
\begin{tabular}{l|lll}
    $n$ & $\leq 13$
        & $14$
        & $15$ \\
    \hline
    $F^{3,1}_n$ & $0$
        & $4$
        & $59$
\end{tabular}
\end{center}
Using the methods of Section~\ref{sec:emb},
we have determined these graphs unembeddable.
\end{comp}
We need one final computational result.
\begin{comp}\label{comp:bic2}
There are no connected graphs
on less than~$15$ vertices
with minimal vertex-order~$2$
with a vertex with fixed color~$0$. 
\end{comp}
We are ready to prove that a minimal Kochen-Specker graph
is biconnected.
\begin{proof}[Proof of Proposition~\ref{prop:ks-biconn}]
Suppose we are given a connected but not biconnected graph~$G$.
Then we can decompose
this graph into its biconnected components.  A biconnected components
is a maximal subgraph that is biconnected.
We can consider the graph of biconnected components~$B(G)$:
two biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the biconnected components in the graph,
would be itself biconnected.  Thus~$B(G)$ is a tree.

Furthermore suppose, reasoning towards contradication,
that~$G$ is a minimal Kochen-Specker graph.
Now consider a leaf~$A$ in the biconnected component graph~$B(G)$.
That is, $A$ is a biconnected component that is connected to only
one other biconnected component.
Let~$B$ be the remainder of the graph.
There is exactly one pair~$(a,b)$ with~$a$ in~$A$ and~$b$ in~$B$
and~$a$ adjacent to~$b$.

@@

Let~$A'$ denote the graph containing two copies of~$A$
connected at the copies of the vertex with fixed color~$1$.
Using the reasoning in the proof of Lemma~\ref{lem:biconn},
we see~$A'$ is a Kochen-Specker graph that is not biconnected.
Define~$B'$ similarly. Observe that~$A'$ and~$B'$ must have the same
number of vertices as~$G$ by the minimality of~$G$.

@@

We can go one step further.
\begin{prop}
    A minimal Kochen-Specker graph is edge-triconnected.
\end{prop}
\begin{proof}
    Given a minimal Kochen-Specker graph~$G$.
    Suppose it is not edge-triconnected.
    Then it splits into two graphs~$A$ and~$B$
    together with verteces~$a_1, a_2 \in A$
    and~$b_1,b_2 \in B$
    such that~$(a_1, b_1)$ and~$(a_2,b_2)$
    are the only edges between~$A$ and~$B$.
    Note that~$A$ and~$B$ must be $010$-colorable,
    for otherwise~$G$ would not be a minimal Kochen-Specker graph.
    \begin{enumerate}
    \item Suppose~$a_1 = a_2$ and~$b_1 = b_2$.
                Then~$G$ is biconnected.  Contradiction with
                    Proposition~\ref{prop:ks-biconn}.
    \item Suppose~$a_1 \neq a_2$ and~$b_1 = b_2$.
        \label{tricase1}
        Suppose~$b_1=b_2$ does not have a fixed color in~$B$.
        Then any coloring of~$A$ can be extended with some coloring
        in~$B$ to a coloring of~$G$.  Contradiction.
        Apparently~$b_1=b_2$ has a fixed color in~$B$.
        \begin{enumerate}
            \item
            Suppose~$b_1=b_2$ has fixed color~$1$ in~$B$.
            Note~$\#B \geq 16$.

@@

Although these restrictions are theoretically pleasing,
they seem to be of little use as a practical restriction.
Concerning excluding unconnected graphs:
% TODO Add argument or computation for {2,3}-connected
% TODO note that restricting to (bi/tri)-connected is a waste of time
% TODO add graph 
\begin{comp}
    There are five non-isomorphic minimal
    squarefree connected graphs
    with minimal vertex order 3 and they have 10 vertices.
\end{comp}
\begin{cor}
    Any unconnected
    squarefree graph with minimal vertex order 3
    has at least 20 vertices, for it has two connected components,
    each with at least 10 vertices.
    With 20 vertices, there are exactly 25 of these.
\end{cor}
This justifies, at this stage, not checking for connectedness.
Similarly, we believe there are very few connected but not biconnected graphs.

>>>>

We continue with a strengthening of Proposition~\ref{prop:ks-conn}.
\begin{prop}\label{prop:ks-biconn}
A minimal Kochen-Specker graph is edge-biconnected,
that is: removing any single edge leaves the graph connected.
\end{prop}
We need some preparation, before we can prove this Proposition.
\begin{dfn}
Given a graph~$G$ and a vertex~$v$ of~$G$.
We say, \keyword{$v$ has fixed color~$c$ (in~$G$)},
if~$G$ is~$010$-colorable
and for every~$010$-coloring of~$G$,
the vertex~$v$ is assigned color~$c$.
\end{dfn}
We are interested in these graphs because of the following observation.
\begin{lem}\label{lem:biconn}
If there is an embeddable graph~$G$ on~$n$ vertices with a vertex
with fixed color~$1$,
then there is a Kochen-Specker graph on~$2n$ vertices.
\end{lem}
\begin{proof}
Let~$G$ be a graph and~$v$ a vertex of~$G$ with fixed color~$1$.
Consider two copies of the graph~$G$.
Connect the two instances of~$v$ with an edge.
Call this graph~$G'$.
Clearly, $G'$ is not~$010$-colorable.

@@

Suppose there are points in~$S'$ and~$S$
that overlap.
Note that the north pole (and south pole) is not in~$S'$. 
Let~$S''$ be~$S'$ rotated along the north pole at some angle~$\alpha$.
There are finitely many angles such that there are overlapping points.
Thus there is an angle such that~$S \cup S''$ is an embedding of~$G'$.
\end{proof}
Unfortunately, these graphs are not small.
\begin{comp}\label{comp:bic1}
Let~$F^{3,1}_n$ denote the number of connected graphs
with minimal vertex-order~$3$
on~$n$ vertices
with a vertex with fixed color~$1$.
Then:
\begin{center}
\begin{tabular}{l|lll}
    $n$ & $\leq 13$
        & $14$
        & $15$ \\
    \hline
    $F^{3,1}_n$ & $0$
        & $4$
        & $59$
\end{tabular}
\end{center}
Using the methods of Section~\ref{sec:emb},
we have determined these graphs unembeddable.
\end{comp}
We need one final computational result.
\begin{comp}\label{comp:bic2}
There are no connected graphs
on less than~$15$ vertices
with minimal vertex-order~$2$
with a vertex with fixed color~$0$. 
\end{comp}
We are ready to prove that a minimal Kochen-Specker graph
is edge-biconnected.
\begin{proof}[Proof of Proposition~\ref{prop:ks-biconn}]
Suppose we are given a connected but not edge-biconnected graph~$G$.
Then we can decompose
this graph into its edge-biconnected components.  A edge-biconnected components
is a maximal subgraph that is edge-biconnected.
We can consider the graph of edge-biconnected components~$B(G)$:
two edge-biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of edge-biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the edge-biconnected components in the graph,
would be itself edge-biconnected.  Thus~$B(G)$ is a tree.

Furthermore suppose, reasoning towards contradication,
that~$G$ is a minimal Kochen-Specker graph.
Now consider a leaf~$A$ in the edge-biconnected component graph~$B(G)$.
That is, $A$ is a edge-biconnected component that is connected to only
one other edge-biconnected component.
Let~$B$ be the remainder of the graph.
There is exactly one pair~$(a,b)$ with~$a$ in~$A$ and~$b$ in~$B$
and~$a$ adjacent to~$b$.

@@

Let~$A'$ denote the graph containing two copies of~$A$
connected at the copies of the vertex with fixed color~$1$.
Using the reasoning in the proof of Lemma~\ref{lem:biconn},
we see~$A'$ is a Kochen-Specker graph that is not edge-biconnected.
Define~$B'$ similarly. Observe that~$A'$ and~$B'$ must have the same
number of vertices as~$G$ by the minimality of~$G$.

@@

We can go one step further.
\begin{prop}
    A minimal Kochen-Specker graph is edge-triconnected.
    That is: removing any two edges keeps the graph connected.
\end{prop}
\begin{proof}
    Given a minimal Kochen-Specker graph~$G$.
    Suppose it is not edge-triconnected.
    Then it splits into two graphs~$A$ and~$B$
    together with verteces~$a_1, a_2 \in A$
    and~$b_1,b_2 \in B$
    such that~$(a_1, b_1)$ and~$(a_2,b_2)$
    are the only edges between~$A$ and~$B$.
    Note that~$A$ and~$B$ must be $010$-colorable,
    for otherwise~$G$ would not be a minimal Kochen-Specker graph.
    \begin{enumerate}
    \item Suppose~$a_1 = a_2$ and~$b_1 = b_2$.
                Then~$G$ is not edge-biconnected.  Contradiction with
                    Proposition~\ref{prop:ks-biconn}.
    \item Suppose~$a_1 \neq a_2$ and~$b_1 = b_2$.
        \label{tricase1}
        Suppose~$b_1=b_2$ does not have a fixed color in~$B$.
        Then any coloring of~$A$ can be extended with some coloring
        in~$B$ to a coloring of~$G$.  Contradiction.
        Apparently~$b_1=b_2$ has a fixed color in~$B$.
        \begin{enumerate}
            \item
            Suppose~$b_1=b_2$ has fixed color~$1$ in~$B$.
            Note~$\#B \geq 16$.

@@

Although these restrictions are theoretically pleasing,
they seem to be of little use as a practical restriction.
Concerning excluding unconnected graphs:
% TODO Add argument or computation for {2,3}-connected
% TODO note that restricting to (bi/tri)-connected is a waste of time
% TODO add graph 
\begin{comp}
    There are five non-isomorphic minimal
    squarefree connected graphs
    with minimal vertex order 3 and they have 10 vertices.
\end{comp}
\begin{cor}
    Any unconnected
    squarefree graph with minimal vertex order 3
    has at least 20 vertices, for it has two connected components,
    each with at least 10 vertices.
    With 20 vertices, there are exactly 25 of these.
\end{cor}
This justifies, at this stage, not checking for connectedness.
Similarly, we believe there are very few connected but not
edge-biconnected graphs.
